---
title: "Report on Lake Ontario Microbes"
author: "Maria Calderon-Marrero"
date: "2026-02-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load packages

First we need to ensure we have access to the functions we need for our analysis.

```{r load-packages}
library(tidyverse)
```

# Goal of this report

Here, I create an analysis of environmental and microbial data from Lake Ontario. The goal is to determine how environmental variables like temperature affect the abundance of different Phyla of bacteria.

# Read in the Data

We need two files: 1)The Taxon File (cleaned up); 2) The Sample Data

## 1)Taxonomy Data

```{r load-data-taxon}
# Load taxon abundance data
taxon_dirty_df <-
  read_csv("data/taxon_abundance.csv", skip=2) %>%
  select(-...10) %>%
  rename(sequencer = ...9)

#Look at the top of the dataframe
head(taxon_dirty_df, 6)
#View(taxon_dirty_df)

# CLean up data by removing columns
taxon_clean_df <-
  taxon_dirty_df %>%
  select(sample_id : Cyanobacteria)

#How many columns? Intuition check!
ncol(taxon_clean_df) #Clean data frame
ncol(taxon_dirty_df) #Dirty data frame
```

## 2) Environmental Data

Let's load second data set

```{r load-data-samples}
sample_data_df <-
  read_csv("data/sample_data.csv")

#Quick Look
head(sample_data_df)

```


# Reshaping the data

Create long formatted data from wide-formatte data

```{r reshape data}
taxon_long_df <-
taxon_clean_df %>%
  pivot_longer(cols = Proteobacteria : Cyanobacteria,
               names_to = "Phylum", values_to = "Abundance")

#Intuition Check
nrow(taxon_long_df)
nrow(taxon_dirty_df)
```


## Plot Taxon Data

```{r plot-taxa}
taxon_long_df %>%
  ggplot(aes(x = sample_id, y = Abundance, fill = Phylum))+
           geom_col() +
           theme(axis.text.x = element_text(angle = 90))
```










